### 2018-03-08 chapter4/ms.tex
Chapter 4 Second Pass. Bulk formatting

A second high-significance GW event, designated GW151226, was
discovered by LIGO on 2015 December 26 at 03:38:53 UT \citep{abb+16c}.
This event was also due to the inspiral and merger of a BBH system, consisting
of $14.2^{+8.3}_{-3.7}$~M$_\odot$ and $7.5^{+2.3}_{-2.3}$~M$_\odot$ black holes at a luminosity
distance of $d_L = 440^{+180}_{-190}$~Mpc \citep{LIGOGW151226}.
The initial localization was provided as a probability sky map via a private
GCN circular 37 hours after the GW detection \citep{SingerPrice16}. We initiated
optical follow-up observations with DECam 10 hours later on 2015 December 28,
and imaged a 28.8 deg$^2$ region in the $i$ and $z$ bands during several epochs. Here we report
the results of this search. In \cref{sec:ch4_obs} we discuss the
observations and data analysis procedures. In \cref{sec:ch4_analysis}
we present our search methodology for potential counterparts to GW151226,
and the results of this search. We summarize our conclusions in \cref{sec:ch4_conc}. We perform
cosmological calculations assuming $H_0 = 67.8$~km  s$^{-1}$ Mpc$^{-1}$,
$\Omega_{\lambda} = 0.69$, and $\Omega_m = 0.31$ \citep{Planck2016}.
 Magnitudes are reported in the AB system.

\section{Observations and Data Reduction}
\label{sec:ch4_obs}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.75 \textwidth]{./figs/chapter4/fig1.pdf}
\caption{
Sky region covered by our DECam observations (red
hexagons) relative to the 50\% and 90\% probability regions
from the {\tt BAYESTAR} (cyan contours) and {\tt LALInference}
(white contours) localization of GW151226.
The background color indicates the estimated $5\sigma$
point-source limiting magnitude for a 90~s $i$-band exposure
as a function of sky position for the first night of our DECam
observations. The variation in the limiting magnitude is largely
driven by the dust extinction and airmass at that position.
The dark grey regions indicate sky positions that
were unobservable due to the telescope pointing limits.  The yellow contour
indicates the region of sky covered by the Dark Energy Survey
(DES). The total effective area for the 12 DECam pointings is 28.8
deg$^2$, corresponding to 3\%~(2\%) of the probability in the
{\tt BAYESTAR}~({\tt LALInference}) sky map.}
\label{fig:ch4_obs}
\end{figure}

We initiated follow-up observations with DECam on 2015 December 28 UT,
two days after the GW detection and 10 hours after distribution of the {\tt BAYESTAR}
sky map. DECam is a wide-field optical imager with a 3 deg$^2$ field of view
\citep{Flaugher+15}. We imaged a 28.8 deg$^2$ region corresponding to 3\%
of the sky localization probability when convolved with the initial {\tt BAYESTAR} map
and 2\% of the localization probability in the final {\tt LALInference} sky map.
The pointings and ordering of the DECam observations were determined using
the automated algorithm described in \citet{GW150914DECam}. The choice
of observing fields was constrained by weather, instrument availability, and the available
time to observe this sky region given its high airmass. We
obtained four epochs of data with each epoch consisting of one 90 s
exposure in $i$-band and two 90 s exposures in $z$-band for each of the 12 pointings. The first
epoch was obtained 2--3 days after the initial LIGO trigger (2015
December 28--29 UT), the second epoch was at 6 days (2016 January 1
UT), the third epoch was at 13--14 days (2016 January 8--9), and the
fourth epoch was at 23--24 days (2016 January 18--19). A summary of
the observations is provided in \cref{datatable} and a visual
representation of the sky region is shown in
\Cref{fig:ch4_obs}.

@@

In the case of the GW151226 observations, we began with raw images
acquired from the NOAO archive\footnote{http://archive.noao.edu/} and the
most recent calibration files\footnote{http://www.ctio.noao.edu/noao/content/decam-calibration-files}.
Astrometric calibration was performed relative to the Pan-STARRS1 (PS1)
$3\pi$ survey and 2MASS $J$-band catalogs. The two $z$-band exposures
were then coadded. Photometric calibration was performed using the
PS1 $3\pi$ survey with appropriate calibrations between
PS1 and DECam magnitudes \citep{Scolnic+15}. Image subtraction was performed
using observations from the final epoch as templates. The approach
to candidate selection is described in \cref{sec:ch4_analysis}.

\begin{deluxetable}{lrccccccc}
\tabletypesize{\footnotesize}
\tablecolumns{9}
\tablewidth{0pt}
\tablecaption{Summary of DECam Observations of GW151226
\label{datatable}}
\tablehead{
\colhead{Visit}    &
\colhead{UT}       &
\colhead{$\Delta t$\tablenotemark{a}} &
\colhead{$\langle$PSF$_i$$\rangle$}    &
\colhead{$\langle$PSF$_z$$\rangle$}    &
\colhead{$\langle$airmass$\rangle$}  &
\colhead{$\langle$depth$_i\rangle$}   &
\colhead{$\langle$depth$_z\rangle$}  &
\colhead{$A_{\mathrm{eff}}$\tablenotemark{b}} \\
\colhead{ } &
\colhead{ } &
\colhead{(days)} &
\colhead{(arcsec)} &
\colhead{(arcsec)} &
\colhead{ } &
\colhead{(mag)} &
\colhead{(mag)} &
\colhead{(deg$^2$)}
}
\startdata
 Epoch 1 & 2015-12-28.11 & 1.96 & 0.97 & 0.99 & 1.95 & 22.39 & 22.23 & 14.4 \\      %  6 hexes
               & 2015-12-29.11 & 2.96 & 1.00 & 0.97 & 1.78 & 22.57 & 22.46 & 14.4 \\      %  6 hexes
                \hline
 Epoch 2 & 2016-01-01.06 & 5.91 & 0.95 & 0.90 & 1.57 & 21.37 & 21.06 & 28.8 \\      % 12 hexes
 \hline
 Epoch 3 & 2016-01-08.11 & 12.96 & 1.68 & 1.62 & 2.15 & 22.09 & 21.70 & 24.0 \\      % 10 hexes
                & 2016-01-09.11 & 13.96 & 1.17 & 1.12 & 1.80 & 22.44 & 22.17 &  4.8 \\      %  2 hexes
                 \hline
 Epoch 4 & 2016-01-18.03 & 22.88 & 1.21 & 1.20 & 1.48 & 22.00 & 22.01 & 12.0 \\      %  5 hexes
               & 2016-01-19.01 & 23.86 & 1.29 & 1.25 & 1.71 & 21.86 & 21.90 & 16.8 \\ %  7 hexes
\enddata
\tablecomments{Summary of our DECam follow-up observations of
GW151226. The PSF, airmass, and depth are the average values across
all observations on that date. The reported depth corresponds to the mean
$5\sigma$ point source detection in the coadded search images.}
\tablenotetext{a}{Time elapsed between the GW trigger time and the time of the first image.}
\tablenotetext{b}{The effective area corresponds to 12 DECam pointings
taking into account that $\approx 20\%$ of the 3
deg$^2$ field of view of DECam is lost due to chip gaps (10\%), 3
dead CCDs (5\%, \citealt{Diehl+14}), and masked edge pixels (5\%).}
\end{deluxetable}

@@

\item We require a $\ge 3\sigma$ decline in flux between the first and
third epochs to search for significant fading\footnote{We note that this
criterion effectively requires the detection in the first epoch to be $\gtrsim 5\sigma$
producing an effectively shallower transient search. Soares-Santos et al.
\citeyearpar{ss+16} quantified this effect by injecting fake sources into
their observations to determine the recovery efficiency and loss of detection
depth from analysis cuts. Here, we forego
such analysis to focus the discussion on the effects of contamination in
optical follow-up of GW events.}. We calculate $\sigma$ as the
quadrature sum of the flux errors from the first and third epochs
($\sigma = \sqrt{\sigma_1^2 + \sigma_3^2}$, where $\sigma_1$ and
$\sigma_3$ are the flux errors from the first and third epochs,
respectively). This criterion leaves a sample of 48 objects.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.95\textwidth]{./figs/chapter4/fig2.pdf}
\caption{{\it Top}: Single-epoch images of our main candidate
from all four epochs (green circle). This is the event discovered as PS15cdi in
the PSST about 94 d prior to GW151226. {\it Bottom}: Light
curve data for PS15cdi from PSST $w$- and $i$-band
observations (green squares and yellow diamonds, respectively).
Our DECam $i$- and $z$-band data are shown as
blue circles and red stars, respectively. The revised DECam analysis
using pre-existing templates is shown as open symbols. Upper
limits are indicated by triangles. The inset focuses on our
DECam data, indicating a rapid decline in both $i$ and $z$
bands.  We fit a power-law model to the data finding a temporal index of
$\alpha = -0.43$ (dashed-dot line). Kilonova models from Barnes and
Kasen \citeyearpar{BarnesKasen13} with $v_{\rm ej}=0.2c$ and $M_{\rm ej}=0.1$ M$_\odot$
at a distance of $187$~Mpc are also shown (dashed line).}
\label{fig:ch4_PS15cdi}
\end{figure}

The final candidate in our search is located at RA = 01$^{\rm h}$42$^{\rm m}$16$^{\rm s}$.17
and DEC = $-$02\arcdeg13\arcmin42 6\arcsec~(J2000), with an offset of $5.8$ arcsec from the galaxy
CGCG 386-030 (RA =  01$^{\rm h}$42$^{\rm m}$15$^{\rm s}$.6, DEC = $-$02\arcdeg13\arcmin38
5\arcsec; J2000), at $z = 0.041$ or $d_L \approx 187$ Mpc (6dFGS, \citealt{Jones+04,Jones+09}); see
\Cref{fig:ch4_PS15cdi}. We note that this distance is inconsistent with the 90\% confidence interval for
the distance to GW151226 based on the GW data \citep{LIGOGW151226}. We observe this source in a state of rapid decline with an
absolute magnitude of $M_i \approx -15$~mag on 2015 December 28 and $M_i\approx -14.5$~mag on 2016
January 1, indicating a decline rate of $\approx 0.12$ mag d$^{-1}$; the decline rate in $z$-band is
$\approx 0.10$ mag d$^{-1}$. Additionally, the source exhibits a red $i-z$ color of $0.3$ mag. We
fit these data to a power-law model typical for GRB afterglows $(F_\nu \propto \nu^{\beta} t^{\alpha})$
and find a temporal index of $\alpha = -0.43\pm0.12$ and a spectral index of $\beta = -1.8\pm0.8$,
both of which differ from the expected values for GRB afterglows ($\alpha \approx -1$,
$\beta \approx -0.75$, \citealt{Sari+98}). Additionally, we compare our observations to a
kilonova model with ejecta parameters of $v_{\rm ej} = 0.2$c and $M_{\rm ej} = 0.1$ M$_\odot$
 \citep{BarnesKasen13}. We find that the timescale of the transient agrees with those expected for kilonovae,
 but the color is bluer than the expected value of $i-z \approx 1$~mag \citep{BarnesKasen13}. Thus, the properties
 of this transient differ from those of GRB afterglows or kilonovae. The observations
 and models are shown in \Cref{fig:ch4_PS15cdi}.

This source was previously detected as PS15cdi on 2015 September 23 by the Pan-STARRS
Survey for Transients (PSST\footnote{\url{http://psweb.mp.qub.ac.uk/ps1threepi/psdb/candidate/1014216170021342600/}},
\citealt{Huber+15}); see \Cref{fig:ch4_PS15cdi}. The absolute $i$-band
magnitude in the first PSST epoch, $M_i\approx -16.6$ mag and the shallow decline of $\approx0.6$~mag
over $\approx70$~d, are consistent with a Type IIP core-collapse supernova (SN).
A likely interpretation of the rapid decline in our observations is that PS15cdi is a Type IIP SN undergoing
the rapid transition from the hydrogen recombination driven plateau to the radioactive $^{56}$Co dominated
phase \citep{KasenWoosley09,Sanders+15,Dhungana+16}. The red $i-z$ color in our data is consistent with observations
of other IIP SN during this phase of evolution (e.g., SN2013ej, \citealt{Dhungana+16}).
This transition typically occurs about 100~d post explosion \citep{KasenWoosley09,Sanders+15,Dhungana+16}, consistent with the
timing of our observations relative to the first detection in PSST.

@@

Clearly, we can rule out this candidate based on the PSST detections prior to
GW151226, but without this crucial information this candidate
would have been a credible optical counterpart based on its light curve behavior and distance.
It is therefore useful to develop an understanding of the expected rates for such contaminants
to inform expectations in future searches. We adopt a local core-collapse SN rate of
$7\times10^{-5}$~yr$^{-1}$~Mpc$^{-3}$ \citep{Li+11,Cappellaro+15},
and a Type IIP SN fraction of 48\% of this rate \citep{Smith+11}. The rapid
decline phase typically lasts about $20$~d \citep{KasenWoosley09,Sanders+15,Dhungana+16},
so we consider events that occur within that time frame. Lastly, given its apparent brightness,
we assume that PS15cdi represents the approximate maximum distance to which we can observe
these events in our data. We thus find an expected occurrence rate of $\sim 0.04$ events in our search area
making our detection of PS15cdi somewhat unlikely, and indicating that $\lesssim 1$ such
events are expected in a typical GW localization region.

>>>>

A second high-significance GW event, designated GW151226, was
discovered by LIGO on 2015 December 26 at 03:38:53 UT \citep{LIGOGW151226}.
This event was also due to the inspiral and merger of a BBH system, consisting
of $14.2^{+8.3}_{-3.7}$~M$_\odot$ and $7.5^{+2.3}_{-2.3}$~M$_\odot$ black holes at a luminosity
distance of $d_L = 440^{+180}_{-190}$~Mpc \citep{LIGOGW151226}.
The initial localization was provided as a probability sky map via a private
GCN circular 37 hours after the GW detection \citep{SingerPrice16}. We initiated
optical follow-up observations with DECam 10 hours later on 2015 December 28,
and imaged a 28.8 deg$^2$ region in the $i$ and $z$ bands during several epochs. Here we report
the results of this search. In \cref{sec:ch4_obs} we discuss the
observations and data analysis procedures. In \cref{sec:ch4_analysis}
we present our search methodology for potential counterparts to GW151226,
and the results of this search. We summarize our conclusions in \cref{sec:ch4_conc}. We perform
cosmological calculations assuming $H_0 = 67.8$~km  s$^{-1}$ Mpc$^{-1}$,
$\Omega_{\lambda} = 0.69$, and $\Omega_m = 0.31$ \citep{Planck2016}.
 Magnitudes are reported in the AB system.

\clearpage
\section{Observations and Data Reduction}
\label{sec:ch4_obs}

@@

We initiated follow-up observations with DECam on 2015 December 28 UT,
two days after the GW detection and 10 hours after distribution of the {\tt BAYESTAR}
sky map. DECam is a wide-field optical imager with a 3 deg$^2$ field of view
\citep{Flaugher+15}. We imaged a 28.8 deg$^2$ region corresponding to 3\%
of the sky localization probability when convolved with the initial {\tt BAYESTAR} map
and 2\% of the localization probability in the final {\tt LALInference} sky map.
The pointings and ordering of the DECam observations were determined using
the automated algorithm described in \citet{GW150914DECam}. The choice
of observing fields was constrained by weather, instrument availability, and the available
time to observe this sky region given its high airmass. We
obtained four epochs of data with each epoch consisting of one 90 s
exposure in $i$-band and two 90 s exposures in $z$-band for each of the 12 pointings. The first
epoch was obtained 2--3 days after the initial LIGO trigger (2015
December 28--29 UT), the second epoch was at 6 days (2016 January 1
UT), the third epoch was at 13--14 days (2016 January 8--9), and the
fourth epoch was at 23--24 days (2016 January 18--19). A summary of
the observations is provided in \cref{ch4_datatable} and a visual
representation of the sky region is shown in
\Cref{fig:ch4_obs}.

\begin{deluxetable}{lrccccccc}
\tabletypesize{\footnotesize}
\tablecolumns{9}
\tablewidth{0pt}
\tablecaption{Summary of DECam Observations of GW151226
\label{ch4_datatable}}
\tablehead{
\colhead{Visit}    &
\colhead{UT}       &
\colhead{$\Delta t$\tablenotemark{a}} &
\colhead{$\langle$PSF$_i$$\rangle$}    &
\colhead{$\langle$PSF$_z$$\rangle$}    &
\colhead{$\langle$airmass$\rangle$}  &
\colhead{$\langle$depth$_i\rangle$}   &
\colhead{$\langle$depth$_z\rangle$}  &
\colhead{$A_{\mathrm{eff}}$\tablenotemark{b}} \\
\colhead{ } &
\colhead{ } &
\colhead{(days)} &
\colhead{(arcsec)} &
\colhead{(arcsec)} &
\colhead{ } &
\colhead{(mag)} &
\colhead{(mag)} &
\colhead{(deg$^2$)}
}
\startdata
 Epoch 1 & 2015-12-28.11 & 1.96 & 0.97 & 0.99 & 1.95 & 22.39 & 22.23 & 14.4 \\      %  6 hexes
               & 2015-12-29.11 & 2.96 & 1.00 & 0.97 & 1.78 & 22.57 & 22.46 & 14.4 \\      %  6 hexes
                \hline
 Epoch 2 & 2016-01-01.06 & 5.91 & 0.95 & 0.90 & 1.57 & 21.37 & 21.06 & 28.8 \\      % 12 hexes
 \hline
 Epoch 3 & 2016-01-08.11 & 12.96 & 1.68 & 1.62 & 2.15 & 22.09 & 21.70 & 24.0 \\      % 10 hexes
                & 2016-01-09.11 & 13.96 & 1.17 & 1.12 & 1.80 & 22.44 & 22.17 &  4.8 \\      %  2 hexes
                 \hline
 Epoch 4 & 2016-01-18.03 & 22.88 & 1.21 & 1.20 & 1.48 & 22.00 & 22.01 & 12.0 \\      %  5 hexes
               & 2016-01-19.01 & 23.86 & 1.29 & 1.25 & 1.71 & 21.86 & 21.90 & 16.8 \\ %  7 hexes
\enddata
\tablecomments{Summary of our DECam follow-up observations of
GW151226. The PSF, airmass, and depth are the average values across
all observations on that date. The reported depth corresponds to the mean
$5\sigma$ point source detection in the coadded search images.}
\tablenotetext{a}{Time elapsed between the GW trigger time and the time of the first image.}
\tablenotetext{b}{The effective area corresponds to 12 DECam pointings
taking into account that $\approx 20\%$ of the 3
deg$^2$ field of view of DECam is lost due to chip gaps (10\%), 3
dead CCDs (5\%, \citealt{Diehl+14}), and masked edge pixels (5\%).}
\end{deluxetable}

\begin{figure}[t!]
\centering
\vspace{-60pt}
\includegraphics[width=0.75 \textwidth]{./figs/chapter4/fig1.pdf}
\vspace{-60pt}
\caption{
Sky region covered by our DECam observations (red
hexagons) relative to the 50\% and 90\% probability regions
from the {\tt BAYESTAR} (cyan contours) and {\tt LALInference}
(white contours) localization of GW151226.
The background color indicates the estimated $5\sigma$
point-source limiting magnitude for a 90~s $i$-band exposure
as a function of sky position for the first night of our DECam
observations. The variation in the limiting magnitude is largely
driven by the dust extinction and airmass at that position.
The dark grey regions indicate sky positions that
were unobservable due to the telescope pointing limits.  The yellow contour
indicates the region of sky covered by the Dark Energy Survey
(DES). The total effective area for the 12 DECam pointings is 28.8
deg$^2$, corresponding to 3\%~(2\%) of the probability in the
{\tt BAYESTAR}~({\tt LALInference}) sky map.}
\label{fig:ch4_obs}
\end{figure}

@@

\clearpage
In the case of the GW151226 observations, we began with raw images
acquired from the NOAO archive\footnote{http://archive.noao.edu/} and the
most recent calibration files\footnote{http://www.ctio.noao.edu/noao/content/decam-calibration-files}.
Astrometric calibration was performed relative to the Pan-STARRS1 (PS1)
$3\pi$ survey and 2MASS $J$-band catalogs. The two $z$-band exposures
were then coadded. Photometric calibration was performed using the
PS1 $3\pi$ survey with appropriate calibrations between
PS1 and DECam magnitudes \citep{Scolnic+15}. Image subtraction was performed
using observations from the final epoch as templates. The approach
to candidate selection is described in \cref{sec:ch4_analysis}.

@@

\clearpage
\item We require a $\ge 3\sigma$ decline in flux between the first and
third epochs to search for significant fading\footnote{We note that this
criterion effectively requires the detection in the first epoch to be $\gtrsim 5\sigma$
producing an effectively shallower transient search. Soares-Santos et al.
\citeyearpar{ss+16} quantified this effect by injecting fake sources into
their observations to determine the recovery efficiency and loss of detection
depth from analysis cuts. Here, we forego
such analysis to focus the discussion on the effects of contamination in
optical follow-up of GW events.}. We calculate $\sigma$ as the
quadrature sum of the flux errors from the first and third epochs
($\sigma = \sqrt{\sigma_1^2 + \sigma_3^2}$, where $\sigma_1$ and
$\sigma_3$ are the flux errors from the first and third epochs,
respectively). This criterion leaves a sample of 48 objects.

@@

\clearpage
The final candidate in our search is located at RA = 01$^{\rm h}$42$^{\rm m}$16$^{\rm s}$.17 and DEC = $-$02\arcdeg13\arcmin42 6\arcsec~(J2000), with an offset of $5.8$ arcsec from the galaxy CGCG 386-030 (RA =  01$^{\rm h}$42$^{\rm m}$15$^{\rm s}$.6, DEC = $-$02\arcdeg13\arcmin38 5\arcsec; J2000), at $z = 0.041$ or $d_L \approx 187$ Mpc (6dFGS, \citealt{Jones+04,Jones+09}); see \Cref{fig:ch4_PS15cdi}. We note that this distance is inconsistent with the 90\% confidence interval for the distance to GW151226 based on the GW data \citep{LIGOGW151226}. We observe this source in a state of rapid decline with an absolute magnitude of $M_i \approx -15$~mag on 2015 December 28 and $M_i\approx -14.5$~mag on 2016 January 1, indicating a decline rate of $\approx 0.12$ mag d$^{-1}$; the decline rate in $z$-band is $\approx 0.10$ mag d$^{-1}$. Additionally, the source exhibits a red $i-z$ color of $0.3$ mag.

We fit these data to a power-law model typical for GRB afterglows $(F_\nu \propto \nu^{\beta} t^{\alpha})$ and find a temporal index of $\alpha = -0.43\pm0.12$ and a spectral index of $\beta = -1.8\pm0.8$, both of which differ from the expected values for GRB afterglows ($\alpha \approx -1$, $\beta \approx -0.75$, \citealt{Sari+98}). Additionally, we compare our observations to a kilonova model with ejecta parameters of $v_{\rm ej} = 0.2$c and $M_{\rm ej} = 0.1$ M$_\odot$ \citep{BarnesKasen13}. We find that the timescale of the transient agrees with those expected for kilonovae, but the color is bluer than the expected value of $i-z \approx 1$~mag \citep{BarnesKasen13}. Thus, the properties of this transient differ from those of GRB afterglows or kilonovae. The observations and models are shown in \Cref{fig:ch4_PS15cdi}.

 \begin{figure}[t!]
 \centering
 \includegraphics[width=0.9\textwidth]{./figs/chapter4/fig2.pdf}
 \caption{\singlespace{\it Top}: Single-epoch images of our main candidate
 from all four epochs (green circle). This is the event discovered as PS15cdi in
 the PSST about 94 d prior to GW151226. {\it Bottom}: Light
 curve data for PS15cdi from PSST $w$- and $i$-band
 observations (green squares and yellow diamonds, respectively).
 Our DECam $i$- and $z$-band data are shown as
 blue circles and red stars, respectively. The revised DECam analysis
 using pre-existing templates is shown as open symbols. Upper
 limits are indicated by triangles. The inset focuses on our
 DECam data, indicating a rapid decline in both $i$ and $z$
 bands.  We fit a power-law model to the data finding a temporal index of
 $\alpha = -0.43$ (dashed-dot line). Kilonova models from Barnes and
 Kasen \citeyearpar{BarnesKasen13} with $v_{\rm ej}=0.2c$ and $M_{\rm ej}=0.1$ M$_\odot$
 at a distance of $187$~Mpc are also shown (dashed line).}
 \label{fig:ch4_PS15cdi}
 \end{figure}

\clearpage
This source was previously detected as PS15cdi on 2015 September 23 by the Pan-STARRS
Survey for Transients (PSST\footnote{\url{http://psweb.mp.qub.ac.uk/ps1threepi/psdb/candidate/1014216170021342600/}},
\citealt{Huber+15}); see \Cref{fig:ch4_PS15cdi}. The absolute $i$-band
magnitude in the first PSST epoch, $M_i\approx -16.6$ mag and the shallow decline of $\approx0.6$~mag
over $\approx70$~d, are consistent with a Type IIP core-collapse supernova (SN).
A likely interpretation of the rapid decline in our observations is that PS15cdi is a Type IIP SN undergoing
the rapid transition from the hydrogen recombination driven plateau to the radioactive $^{56}$Co dominated
phase \citep{KasenWoosley09,Sanders+15,Dhungana+16}. The red $i-z$ color in our data is consistent with observations
of other IIP SN during this phase of evolution (e.g., SN2013ej, \citealt{Dhungana+16}).
This transition typically occurs about 100~d post explosion \citep{KasenWoosley09,Sanders+15,Dhungana+16}, consistent with the
timing of our observations relative to the first detection in PSST.

@@

\clearpage
Clearly, we can rule out this candidate based on the PSST detections prior to
GW151226, but without this crucial information this candidate
would have been a credible optical counterpart based on its light curve behavior and distance.
It is therefore useful to develop an understanding of the expected rates for such contaminants
to inform expectations in future searches. We adopt a local core-collapse SN rate of
$7\times10^{-5}$~yr$^{-1}$~Mpc$^{-3}$ \citep{Li+11,Cappellaro+15},
and a Type IIP SN fraction of 48\% of this rate \citep{Smith+11}. The rapid
decline phase typically lasts about $20$~d \citep{KasenWoosley09,Sanders+15,Dhungana+16},
so we consider events that occur within that time frame. Lastly, given its apparent brightness,
we assume that PS15cdi represents the approximate maximum distance to which we can observe
these events in our data. We thus find an expected occurrence rate of $\sim 0.04$ events in our search area
making our detection of PS15cdi somewhat unlikely, and indicating that $\lesssim 1$ such
events are expected in a typical GW localization region.
